Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-07 main.tex

The Axelrod library \cite{knight2016open} is an open source software for
conducting iterated prisoner's dilemma (IPD) research with reproducibilty as a
principal goal. Written in the Python programming language, to date over the
library contains source code contibuted by over 50 individuals from a variety
of geographic locations and technical backgrounds. The library is supported by
a comprehensive test suite that covers all the intended behaviors of the
strategies in the library, as well as the features that conduct matches,
tournaments, and population dynamics.

As of version 3.0.0, the library contains over 200 strategies,
many from the scientific literature, including classic strategies like Win Stay
Lose Shift \cite{nowak1993strategy} and previous tournament winners such as
OmegaTFT \cite{slany2007some}, Adaptive Pavlov \cite{li2007design}, and
ZDGTFT2 \cite{stewart2012extortion}.

@@

In this work we utilize the collection of strategies in the Axelrod library
to train new strategies specifically to win IPD tournaments. We train these
strategies using generic strategy archetypes based on e.g. finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, ashlock2015multiple, ashlock2006changes,
      ashlock2014shaped, ashlock2014evolution, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{gaudesi2016exploiting} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature. Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented and documented. Moreover, as of the time of writing,
we claim that this work contains the best known strategies for the iterated
prisoner's dilemma.

@@

We conducted a tournament with a large collection of strategies from the Axelrod
library, including some additional parametrized strategies (e.g. various parameter
choices for Generous Tit For Tat). These are listed in
Appendix~\ref{app:list_of_players}.
The top 11 performing strategies by median payoff are all strategies trained to maximize
% TODO Potentially modify this as more repetitions are analysed.
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is Desired Belief Strategy (DBS) \cite{Au2006}.
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once, a grudging strategy that defects indefinitely on
the second defection, and Omega Tit For Tat \cite{kendall2007iterated}.
All strategies in the tournament follow a simple set of
rules in accordance with earlier tournaments:

@@

Using the a numerical method of fingerprinting based on
\cite{ashlock2008fingerprinting, ashlock2009fingerprint} we can compare
strategies. The fingerprints of the top performing standard strategies are shown
in Figures~\ref{fig:comparison_fingerprint_standard}. There is a striking
similarity in the fingerprints.

@@

We trained a variety of strategies using evolutionary algorithms, and in the
case of PSO Gambler using a particle swarm algorithm. The evolutionary algorithms
used standard techniques, varying strategies by mutation and crossover, and
evaluating the performance against each opponent for many repetitions. The
best performing strategies in each generation are persisted, variants created,
and objective functions computed again. This process continues for approximately
200 generations or until strategies no longer improve significantly.
% TODO Improve the explanation here with more details. As described this is not
% reproducible.
% TODO Include references

% TODO: Diagram with ANN training performance?

All training code is available on github. There are objective functions for
% TODO Include pointer when code is archived.

\begin{itemize}
 \item total or mean payoff
 \item total or mean payoff difference (unused in this work)
 \item total Moran process wins (fixation probability)
\end{itemize}

These objectives can be easily modified to suit other purposes. New strategies
can be easily trained with variations including noise, spatial structure, and
probabilistically ending matches.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategies did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

@@

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0) that did not include DBS
% Cite v2.2.0
and several other opponents. The precise parameters that are optimal will depend
on the pool of opponents. Moreover we have not extensively trained strategies to
determine the minimum parameters that are sufficient -- neural networks with
fewer nodes and features and finite state machines with fewer states may suffice.
See \cite{ashlock2013impact} for discussion of resource availability for IPD
strategies. It may be possible to train strategies more effective in noisy tournaments
than DBS.

@@

The players used for this study are from Axelrod version 2.13.0. The starred $*$
strategies indicate those that have been trained using reinforcement learning
algorithms.

>>>>

The Axelrod library \cite{axelrodproject, knight2016open} is an open source
software for conducting iterated prisoner's dilemma (IPD) research with
reproducibility as a principal goal. Written in the Python programming language,
to date over the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

The library is continuously developed and as of version 3.0.0, the library
contains over 200 strategies, many from the scientific literature, including
classic strategies like Win Stay Lose Shift \cite{nowak1993strategy} and
previous tournament winners such as OmegaTFT \cite{slany2007some}, Adaptive
Pavlov \cite{li2007design}, and ZDGTFT2 \cite{stewart2012extortion}.

@@

In this work we describe how collections of strategies in the Axelrod library
have been used to train new strategies specifically to win IPD tournaments.
These strategies are trained using generic strategy archetypes based on e.g.
finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, ashlock2015multiple, ashlock2006changes,
      ashlock2014shaped, ashlock2014evolution, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{gaudesi2016exploiting} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature. Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented, tested and documented.
Moreover, as of the time of writing,
we claim that this work contains the best known strategies for the iterated
prisoner's dilemma.

@@

We conducted a tournament with a large collection of strategies from the Axelrod
library, including some additional parametrized strategies (e.g.\ various parameter
choices for Generous Tit For Tat). These are listed in
Appendix~\ref{app:list_of_players}.
The top 11 performing strategies by median payoff are all strategies trained to maximize
% TODO Potentially modify this as more repetitions are analysed.
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is Desired Belief Strategy (DBS) \cite{Au2006}.
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once, a grudging strategy that defects indefinitely on
the second defection, and Omega Tit For Tat \cite{kendall2007iterated}.
All strategies in the tournament follow a simple set of
rules in accordance with earlier tournaments:

@@

Using a numerical method of fingerprinting based on
\cite{ashlock2008fingerprinting, ashlock2009fingerprint} we can compare
strategies. The fingerprints of the top performing standard strategies are shown
in Figures~\ref{fig:comparison_fingerprint_standard}. There is a striking
similarity in the fingerprints.

@@

The trained strategies (denoted by a \(^{*}\) in
Appendix~\ref{app:list_of_players} were trained using reinforcement
learning algorithms. The ideas of reinforcement learning can be attributed to
the original work of \cite{turing1950computing} in which the notion that
computers would learn by taking random actions but according to a distribution
that picked actions with high rewards more often. The two particular algorithms
used here:

\begin{itemize}
    \item Evolutionary algorithm: \cite{moriarty1999evolutionary}.
    \item Particle Swarm Algorithm: \cite{imran2013overview}.
\end{itemize}

The Particle Swarm Algorithm is implemented using the Python library:
\url{https://pypi.python.org/pypi/pyswarm}. This algorithm was used only to
train the Gambler archetype.

All other strategies were trained using evolutionary algorithms. The
evolutionary algorithms used standard techniques, varying strategies by
mutation and crossover, and evaluating the performance against each opponent
for many repetitions. The best performing strategies in each generation are
persisted, variants created, and objective functions computed again.

The default parameters for this procedure:

\begin{itemize}
	\item A population size of 40 individuals (kept constant across the
        generations);
	\item A mutation rate of 10\%;
	\item 10 individuals kept from one generation to the next;
    \item A total of 500 generations.
\end{itemize}

All implementations of these algorithms are archived at
\cite{marc_harper_2017_824264}. This software is (similarly to the Axelrod
library) available on github
\url{https://github.com/Axelrod-Python/axelrod-dojo}.  There are objective
functions for:

\begin{itemize}
 \item total or mean payoff
 \item total or mean payoff difference (unused in this work)
 \item total Moran process wins (fixation probability). This lead to the
     strategies named TF1, TF2, TF3 listed in
     Appendix~\ref{app:list_of_players}.
\end{itemize}

These can be used in Noisy or Standard environments (as evidenced by
Sections~\ref{sec:standard} and~\ref{sec:noisy}). These objectives can be
further modified to suit other purposes. New strategies could be trained with
variations including spatial structure, and probabilistically ending matches.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategies did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
% TODO Question: is this arguably just as much human involvement in the design
% of the lookup tables etc? We could use a lookup table that also picks out the
% action from the middle round, however we have chosen not too.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

@@

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0: \cite{axelrodproject2.2})
that did not include DBS and several other opponents. The precise parameters
that are optimal will depend on the pool of opponents. Moreover we have not
extensively trained strategies to determine the minimum parameters that are
sufficient -- neural networks with fewer nodes and features and finite state
machines with fewer states may suffice. See \cite{ashlock2013impact} for
discussion of resource availability for IPD strategies. It may be possible to
train strategies more effective in noisy tournaments than DBS.

@@

The players used for this study are from Axelrod version 2.13.0
\cite{axelrodproject}. The starred $*$
strategies indicate those that have been trained using reinforcement learning
algorithms.
